### 2019-05-04 papers/haskell19/1-intro.tex
Section 2
MergeMoved stuff from section 2 to intro, work on abstract and conclusions

Pellentesque dapibus suscipit ligula. Donec posuere augue in quam. Etiam vel
tortor sodales tellus ultricies commodo. Suspendisse potenti. Aenean in sem ac
leo mollis blandit. Donec neque quam, dignissim in, mollis nec, sagittis eu,
wisi. Phasellus lacus. Etiam laoreet quam sed arcu. Phasellus at dui in ligula
mollis ultricies. Integer placerat tristique nisl. Praesent augue. Fusce
commodo. Vestibulum convallis, lorem a tempus semper, dui dui euismod elit,
vitae placerat urna tortor vitae lacus. Nullam libero mauris, consequat quis,
varius et, dictum id, arcu. Mauris mollis tincidunt felis. Aliquam feugiat
tellus ut neque. Nulla facilisis, risus a rhoncus fermentum, tellus tellus
lacinia purus, et dictum nunc justo sit amet elit.

The main contribution of this paper are:
%
\begin{itemize}
\item We identify some of the most important limitations of performing random
  testing with automatically derived generators using state-of-the-art tools
  (Section \ref{sec:sources}).
\item We present an extensible mechanism for representing random values built
  upon different source constructions using Data Types \`a la Carte (Section
  \ref{sec:representation}).
\item We develop a modular generation schema, exploiting the composable nature
  of our representations, which are extended to encode information relevant to
  the generation process at the type level (Section \ref{sec:generators}).
\item We present a very simple yet powerful domain specific language for
  describing extensible generators solely based on the types used to represent
  the desired shape of our random data (Section \ref{sec:typelevel}).
\item We provide a Template Haskell tool for automatically deriving all the
  required machinery presented throughout this paper, and evaluate its
  generation performance using three real-world case studies (Section
  \ref{sec:casestudies}).
\end{itemize}

Overall, we present a novel technique for reusing automatically derived
generators in a composable fashion, in contrast with the usual paradigm of
automatically deriving ``closed'' monolithic generators.

>>>>

% --------------------------------------
% Intro
%
Random property-based testing is a quite powerful technique for finding bugs
\cite{ArtsHNS15, HughesPAN16, HughesNSA16}.
%
In the Haskell comunity, QuickCheck is the predominant tool of this such.
%
The developers specify
%
\begin{inparaenum}
\item the testing properties their systems must fullfil, and
\item random data generators for the data types involved at their properties.
\end{inparaenum}
%
Then, QuickCheck generates random seeds using these generators, and uses them to
evaluate the system's properties in search of possible counterexamples, which
often indicate bugs in our system.

% --------------------------------------
% Motivating Example
%
Although QuickCheck provides default random generators for the common base types
like |Int| or |String|, it requires implementing random generators for any
user-derined data type we want to randomly generate.
%
% This section introduces the common workflow for generating random values using
% automatically derived QuickCheck generators, along with its common limitations.
%
To illustrate this, let us consider the following data type definition encoding
basic HTML pages:

\begin{code}
data Html
  =  Text  String
  |  Sing  String
  |  Tag   String Html
  |  Join  Html Html
\end{code}

This type allows us to build HTML pages via four possible data constructors:
|Text| is used for plain text values; |Sing| and |Tag| represent singular and
paired HTML tags, respectively; and |Join| to simply concatenate two HTML pages
one after another.
%
Note that the constructors |Tag| and |Join| are recursive, as they have at least
one field of their own type.
%
In this paper, and in the scope of random generation in general, we will refer
to any non-recursive data constructor as \emph{terminal}, and
\emph{non-terminal} in any other case.

Then, using this representation, the example page\linebreak
%
\texttt{<html><b>hello<br>world!</b></html>}
%
can be encoded simply as:
%
\begin{code}
Tag "html" $ Tag "b" $
  Text "hello" iJoin Sing "br" iJoin Text "world!"
\end{code} %$

% --------------------------------------
% Random Generators
%
Then, in order to generate random values of type |Html|, we need to provide a
random data generator for it.
%
For this purpose, QuickCheck defines the |Arbitrary| type-class for default
defining random generators as:

\begin{code}
class Arbitrary (a :: *) where
  arbitrary :: Gen a
\end{code}
%
where the overloaded symbol |arbitrary :: Gen a| denotes a monadic generator for
values of type |a|.
%
Using this mechanism, we can define a sensible random generator for our |Html|
data type as follows:

\begin{code}
instance Arbitrary Html where
  arbitrary = sized gen
    where  gen 0 = frequency
             [  (2,  Text    <$> arbitrary)
             ,  (1,  Sing    <$> arbitrary) ]
           gen d = frequency
             [  (2,  Text    <$> arbitrary)
             ,  (1,  Sing    <$> arbitrary)
             ,  (4,  Tag     <$> arbitrary  <*> gen (d-1))
             ,  (3,  Join    <$> gen (d-1)  <*> gen (d-1)) ]
\end{code} %$
%
At the top level, this definition parametrizes the generation process using
QuickCheck's |sized| combinator, which lets us build our generator using an
auxliary local function |gen :: Int -> Gen Html|.
%
The |Int| passed to |gen| is known as the \emph{generation size}, and is
threaded seamlessly by QuickCheck on each call to arbitrary.
%
Then, we use this parameter to limit the maximum amount of recursive calls that
our generator can perform, and thus the maximum depth of the generated values.
%
If the generation size is positive (case |gen d|), our random generator picks a
random constructor from our |Html| data type with a given generation frequency
(using QuickCheck's |frequency| combinator), and proceeds to fill its fields
using randomly generated sub-terms---here using Haskell's applicative notation
\cite{Mcbride2008} and the default |Arbitrary| instance for |String|s.
%
For the case of the recursive sub-terms, this generator simply calls itself
recursively with a smaller generation size (|gen (d-1)|).
%
This process repeats until we reach the base case (|gen 0|) on each recursive
sub-term.
%
At this point, our random generator is limited to only pick among the |Html|
terminal constructors, hence terminating the generation process.

% --------------------------------------
% Automatic derivation tools
%
As one can observe, the previous definition is quite mechanical, and depends
only on the generation frequencies we choose for each constructor.
%
Fortunately, there exists a variety of meta-programming tools for obtaining
random generators automatically based on their types.
%
Tools like \emph{MegaDeTH} or \emph{DRAGEN} will synthesize random generators
using the same procedure as the one shown above.
%
\emph{MegaDeTH} simply picks among data constructors uniformingly, whereas
\emph{DRAGEN} analyzes the type definitions and tunes the generation frequencies
based on a simulation-based optimization approach, in order to obtain random
distributions of values based on the users' desires.

% --------------------------------------
% Limitation
%
Such tools provide a useful mechanism for automating the boring and error-prone
task of writing random generators by hand.
%
However, this simple derivation procedure, which only considers the structural
information encoded in data type definitions, and generates one single
constructor at a time; is often too generic to synthesize useful generators in
some practical scenarios.
%
For instance, because of the types we use for deriving our random generators are
too general, leaving a large room for ill-formed values, e.g., invalid HTML
pages; or alternatively, because our properties need to be tested against very
specific patterns of input cases in order to trigger bugs.

This problem, however, can be alleviated by considering additional sources of
structural information from our codebase other than just data constructors.
%
For instance, having a set of high-level combinators, defining an abstract
interface of our data types, e.g., |hr :: Html| or |bold :: Html -> Html|; it is
possible to include them into the generation process, thus obtaining more
complex and structed values.
%
In the same way, having functions defined using pattern matching over the data
types of our interest, e.g. |f (Join (Text t) (Tag s x)) = dots|, it is also
possible to include this rich structure which exposes a relationship among
different data constructors into our random generators.
%
This idea is addressed in depth in the next section, while the rest of this work
focuses on implementing it in a compositional way.

%
% Contributions
%
The main contribution of this paper are:
%
\begin{itemize}
\item We identify some important limitations of performing random testing with
  automatically derived generators using state-of-the-art tools (Section
  \ref{sec:sources}).
\item We present an extensible mechanism for representing random values built
  upon different source constructions using Data Types \`a la Carte (Section
  \ref{sec:representation}).
\item We develop a modular generation schema, exploiting the composable nature
  of our representations, which are extended to encode information relevant to
  the generation process at the type level (Section \ref{sec:generators}).
\item We present a very simple yet powerful domain specific language for
  describing extensible generators solely based on the types used to represent
  the desired shape of our random data (Section \ref{sec:typelevel}).
\item We provide a Template Haskell tool for automatically deriving all the
  required machinery presented throughout this paper, and evaluate its
  generation performance using three real-world case studies (Section
  \ref{sec:casestudies}).
\end{itemize}

Overall, we present a novel technique for reusing automatically derived
generators in a composable fashion, in contrast with the usual paradigm of
automatically deriving ``rigid'', monolithic generators.

### 2019-05-04 papers/haskell19/2-sources.tex
Work on case studies/related work
MergeMoved stuff from section 2 to intro, work on abstract and conclusions

This section introduces the common workflow for generating random values using
automatically derived QuickCheck generators, along with its common limitations.
%
To illustrate this, let us consider the following data type definition encoding
basic HTML pages:

\begin{code}
data Html
  =  Text  String
  |  Sing  String
  |  Tag   String Html
  |  Join  Html Html
\end{code}

This type allows us to build HTML pages via four possible data constructors:
|Text| is used for plain text values; |Sing| and |Tag| represent singular and
paired HTML tags, respectively; and |Join| to simply concatenate two HTML pages
one after another.
%
Note that the constructors |Tag| and |Join| are recursive, as they have at least
one field of their own type.
%
In this paper, and in the scope of random generation in general, we will refer
to any non-recursive data constructor as \emph{terminal}, and
\emph{non-terminal} in any other case.

Then, using this representation the example page\linebreak
%
\texttt{<html>hello<hr><b>world!</b></html>}
%
can be encoded simply as:
%
\begin{code}
Tag "html" $  Text "hello"
              iJoin Single "hr"
              iJoin Tag "b" $ Text "world!"
\end{code} %$

Then, in order to generate random values of type |Html|, QuickCheck requires us
to provide a random data generator for it.

\begin{code}
class Arbitrary (a :: *) where
  arbitrary :: Gen a
\end{code}

\begin{code}
class Arbitrary (a :: *) where
  arbitrary :: Gen a
\end{code}

\begin{code}
instance Arbitrary Html where
  arbitrary = sized gen
    where
      gen 0 = frequency
        [  (2,  Text    <$> arbitrary)
        ,  (1,  Sing    <$> arbitrary) ]
      gen d = frequency
        [  (2,  Text    <$> arbitrary)
        ,  (1,  Sing    <$> arbitrary)
        ,  (4,  Tag     <$> arbitrary  <*> gen (d-1))
        ,  (3,  Join    <$> gen (d-1)  <*> gen (d-1)) ]
\end{code} %$

@@

(<+>) :: Html -> Html -> Html
(<+>) x y = Join x (Join br y)
\end{code} %$

\begin{code}
  gen d = frequency
    [ ...
    , (1, pure br)
    , (5, bold   <$> gen (d-1))
    , (2, list   <$> listOf (gen (d-1)))
    , (3, (<+>)  <$> gen (d-1) <*> gen (d-1)) ]
\end{code}%$

\begin{code}
simplify :: Html -> Html
simplify (Text t1 iJoin Text t2)
  = Text (t1 ++ t2)
simplify ((Text t1 iJoin x) iJoin y)
  = simplify (Text t1 iJoin simplify (x iJoin y))
simplify (x iJoin y)
  = simplify x iJoin simplify y
simplify (Tag t x)
  = Tag t (simplify x)
simplify x = x
\end{code}

@@

>>>>

In this section, we briefly introduce the motivation for including extra
information into our random generators, and discuss about how this can be
implemented in practice.

% --------------------------------------
% Abstract interfaces
%
\subsection{Extracting Information From Interfaces}

A common choice when implementing abstract data types is to transfer the
resposiblity of preserving their invariants to the functions on their abstract
interface.
%
Take for example our |Html| data type.
%
Instead of defining a different data constructor for each possible HTML tag, we
opted for a very small generic representation that can be extended via a set of
combinators making use of this generic interface:

@@

(<+>) :: Html -> Html -> Html
(<+>) x y = Join x (Join br y)
\end{code} %$
%
These functions let us create interesting structured values representing HTML
constructions like linebreaks (|br|), bold blocks (|bold|), unordered lists
(|list|) and concatenation of values one below another (|<+>|).

Note how difficult would be to generate random values containing, for example,
structurally valid HTML lists if we only consider the structural information
encoded in our |Html| data type definion.
%
After all, much of the valid structure of HTML has been encoded on its abstract
interface.

Then, an automatically derived generator could integrate this structural
information into the generation process by creating random values arising from
applying such functions to randomly generated inputs:

\begin{code}
  gen d = frequency
    [ ...
    , (1, pure br)
    , (5, bold   <$> gen (d-1))
    , (2, list   <$> listOf (gen (d-1)))
    , (3, (<+>)  <$> gen (d-1) <*> gen (d-1)) ]
\end{code}%$
%
where the symbol (|...|) represents the rest of the possible sub-generators for
each data constructor seen before.
%
From now on, we will refer to each possible choice given to |frequency| as a
different \emph{random construction}, as we are not considering generating only
ssingle data constructors anymore.

% --------------------------------------
% Functions' pattern matchings
%
\subsection{Extracting Information From Patterns}

A different challenge appears when we try to test functions comprising complex
pattern matchings. Consider for instance the following function for simplifying
|Html| values:

\begin{code}
simplify :: Html -> Html
simplify (Text t1 iJoin Text t2) = Text (t1 ++ t2)
simplify ((Text t1 iJoin x) iJoin y)
  = simplify (Text t1 iJoin simplify (x iJoin y))
simplify (x iJoin y) = simplify x iJoin simplify y
simplify (Tag t x) = Tag t (simplify x)
simplify x = x
\end{code}
%
This function simply traverses |Html| values, joining together every contiguous
pair of |Text| constructors.
%
Ideally, we would like to put approximately the same testing effort info each
clause of |simplify|, or perhaps even more to first two ones, since those are
the ones performing actual simplifications.
%
However, these two clauses are the most difficult ones to test in practice.
%
The probabilty of generating a random value satisfying nested patterns decreases
multiplicatively with the number of constructors we match against to
simultaneously.

In our tests, we were not able to exercise any of these two patterns more than
6\% of the overall testing time (and considering recursive sub-calls), using
random generators derived using either \emph{MegaDeTH} and \emph{DRAGEN}.
%
As expected, most of the random test cases were exercising the simplest (an
rather uninteresting) patterns of this function.

To solve this issue, we could opt to consider each complex pattern as a new kind
of random construction.
%
In this light, we can simply generate values satisfying pattern directly by
returning its corresponding expression, where each variable or wildcard pattern
is filled using a random sub-expression:

@@

% --------------------------------------
% Wrapping up
%
In general, we think that including this static information in our genrators
directly is far from ideal, as in many cases we would like to select which
random constructions are of our interest when testing different properties or
sub-systems.
%
In this light, having ``monolithic'' random generators derived automatically
would complicate our desired workflow.
%
We would need to derive different concrete random generators for each use case
instead of having a single |Arbitrary| instance that can be used everywhere.

In contrast, in this work we opted for developing an extensible fashion, where
the generation machinery is automatically derived only once, and every variant
of our random generators is expressed at the type-level.
%
The following sections of this paper cover our alternative approach in depth.

### 2019-05-04 papers/haskell19/3-representation.tex
MergeMoved stuff from section 2 to intro, work on abstract and conclusions

\begin{code}
instance (Algebra f a, Algebra g a) => Algebra (f oplus'' g) a
  where  alg (InL fa) = alg fa
         alg (InR ga) = alg ga
\end{code}

where we simply use the appropriate |Algebra| instance for the inner
representation, based on which injection was used to construct the composite
value.
%
It is worth remarking that multiple compositions are handled transparently,
i.e., for every set of types |f1|, |f2|, ..., |fk| representing a target type
|t| and having suitable |Algebra| type instances, then the composition |f1
oplus'' f2 oplus'' dots oplus'' fk| can be mapped back to |t| using this same
mechanism.
%
Additionally, the order in which we associate each operand of |oplus| results
semantically irrelevant, thus we can omit parentheses when describing composite
representations.
%
However, we found that in practice, the way in which we decide to associate our
constructions affects the generation speed dramatically.
%
This phenomena is addressed in detail in Section \ref{sec:casestudies}.
%
For simplicity here, we will assume |oplus| as having rigth associativity, i.e.,
|f oplus'' g oplus'' h == f oplus'' (g oplus'' h)|.

@@

\begin{code}
  x' ::  Fix Html'
  x' =   FixC  (InR  (InR (InR (Mk_Join
               (FixC (InL (Mk_Text "hello")))
               (FixC (InR (InL (Mk_Sing "hr"))))))))
\end{code}
%
Where the sequences of |InL| and |InR| data constructors inject each value from
an individual representation into the appropriate position of our composite
representation |Html'|.

@@

\begin{code}
data Pat_simplify_1 r = Mk_simplify_1 String String
data Pat_simplify_2 r = Mk_simplify_2 String r r
\end{code}
%
Where the number after the |#| simply distinguishes the different patterns from
the function |simplify| using the index of the clause they belong to.
%
Once again, we abstracted away every recursive field (this time corresponding to
a recursive pattern variable or wildcard) using a type variable |r| until we
have decided the overall shape of our target data.

>>>>

\begin{code}
instance (Algebra f a, Algebra g a) => Algebra (f oplus'' g) a
  where  alg (InL fa) = alg fa
         alg (InR ga) = alg ga
\end{code}
%
where we simply use the appropriate |Algebra| instance for the inner
representation, based on which injection was used to construct the composite
value.
%
It is worth remarking that multiple compositions are handled transparently,
i.e., for every set of types |f1|, |f2|, ..., |fk| representing a target type
|t| and having suitable |Algebra| type instances, then the composition |f1
oplus'' f2 oplus'' dots oplus'' fk| can be mapped back to |t| using this same
mechanism.
%
Additionally, the order in which we associate each operand of |oplus| results
semantically irrelevant, thus we can omit parentheses when describing composite
representations.
%
However, in practice, the way in which we associate our constructions affects
the generation speed dramatically.
%
This phenomena is addressed in detail in Section \ref{sec:casestudies}.
%
For simplicity here, we will assume |oplus| as having right associativity, i.e.,
|f oplus'' g oplus'' h == f oplus'' (g oplus'' h)|.

@@

\begin{code}
  x' ::  Fix Html'
  x' =   FixC  (InR  (InR (InR (Mk_Join
               (FixC (InL (Mk_Text "hello")))
               (FixC (InR (InL (Mk_Sing "hr"))))))))
\end{code}
%
where the sequences of |InL| and |InR| data constructors inject each value from
an individual representation into the appropriate position of our composite
representation |Html'|.

@@

\begin{code}
data Pat_simplify_1 r = Mk_simplify_1 String String
data Pat_simplify_2 r = Mk_simplify_2 String r r
\end{code}
%
where the number after the |#| simply distinguishes the different patterns from
the function |simplify| using the index of the clause they belong to.
%
Once again, we abstracted away every recursive field (this time corresponding to
a recursive pattern variable or wildcard) using a type variable |r| until we
have decided the overall shape of our target data.

### 2019-05-04 papers/haskell19/8-conclusions.tex
Work on case studies/related work
MergeMoved stuff from section 2 to intro, work on abstract and conclusions

In the future, something something...

>>>>

In the future, we aim to minimize the runtime overhead of our technique, using
perhaps prefix-free encodings based on generation frequencies, as a way to
reduce the representation bureacracy as much as possible.
%
Additionally, we are interested on including additional sources of structural
infrmation from users' codebases, e.g., locally defined patterns or sequencies
of functions applications.
%
Interesting thoughts to motivate our future reseach.

### 2019-05-04 papers/haskell19/abstract.tex
MergeMoved stuff from section 2 to intro, work on abstract and conclusions

Pellentesque dapibus suscipit ligula. Donec posuere augue in quam. Etiam vel
tortor sodales tellus ultricies commodo. Suspendisse potenti. Aenean in sem ac
leo mollis blandit. Donec neque quam, dignissim in, mollis nec, sagittis eu,
wisi. Phasellus lacus. Etiam laoreet quam sed arcu. Phasellus at dui in ligula
mollis ultricies. Integer placerat tristique nisl. Praesent augue. Fusce
commodo. Vestibulum convallis, lorem a tempus semper, dui dui euismod elit,
tellus ut neque. Nulla facilisis, risus a rhoncus fermentum, tellus tellus
lacinia purus, et dictum nunc justo sit amet elit.

Lorem ipsum dolor sit amet, consectetuer adipiscing elit. Donec hendrerit tempor
tellus. Donec pretium posuere tellus. Proin quam nisl, tincidunt et, mattis
eget, convallis nec, purus. Cum sociis natoque penatibus et magnis dis
parturient montes, nascetur ridiculus mus. Nulla posuere. Donec vitae dolor.
Nullam tristique diam non turpis. Cras placerat accumsan nulla. Nullam rutrum.
Nam vestibulum accumsan nisl.

>>>>

Generating ``good'' random values described by algebraic data types is often
quite intricate.
%
State-of-the-art tools for synthesizing random generators serve the valuable
purpose of helping with this task, while providing different level of invariants
imposed over the generated values.
%
However, none of them are built neither for composability nor extendability, a
useful feature when the shape of our random data makes a difference while
testing different properties or sub-systems.
%
% Instead, users are forced to decide wether to manually write their random
% generators or to use rigid automatically derived ones.

In this work, we develop an extensible framework for deriving compositional
generators, which can be easily combined in diferent ways in order to fit the
developers' demands, using a simple type-level domain-specific language.
%
Our framework is based on familiar ideas from the \`a la Carte technique for
writing composable interpreters in Haskell, and extended with the required
machinery for it to work in the scope of random generation of values.
%
Additonally, we provide an implementation of our ideas, and evaluate its
performance using real-world examples.
